An Open and Continuously Updated Fern Tree of Life (FTOL)

### 2022-02-23 ms/manuscript.Rmd
Update MS

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    sanger_sampling,
    coverage_by_rank,
    total_coverage,
    bb_coverage_by_rank,
    bb_total_coverage,
    gb_species_by_year,
    ref_files,
    other_dates,
    crown_ages,
    fossil_ferns_raw,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen
  ),
  store = here::here("_targets")
)
```

@@

We manually inspected any fuzzy matched or non-matching names.
This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
We thus updated and edited the initial World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.

@@

We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].

@@

We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
We selected `r n_calibrations_ferns` fern fossils to use as constraints from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil). 
We assigned fossils to lineages only after consulting the original publication (as opposed to simply relying on the fossil name) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained for a given fossil could change.
We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a.
All constraints other than the root are minimum ages.

@@

The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
Code used to generate FTOL and compile this manuscript is available at https://github.com/pteridophylo/ftol_flow and https://github.com/pteridophylo/ftol_ms, respectively.
A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.

@@

# Count number of raw accessions in final Sanger selection
n_accs_raw_final <- count_accs(sanger_accessions_selection_long)

# Count number of accessions in final Sanger selection, split by locus
n_accs_final <- 
  sanger_accessions_selection_long %>%
  # Count accessions containing different loci as distinct
  mutate(
    name = str_remove_all(name, "accession_"),
    accession = paste(name, accession, sep = "_")) %>%
  count_accs()

n_species_sanger <- 
  sanger_accessions_selection %>% 
  # Remove outgroup and plastome species
  anti_join(
    unique(select(plastome_seqs_combined_filtered, species)),
    by = "species") %>%
  pull(species) %>%
  n_distinct() %>%
  number()

# Analyze join by method
join_by_count <-
  sanger_accessions_selection %>% 
  # Remove outgroup and plastome species
  anti_join(
    unique(select(plastome_seqs_combined_filtered, species)),
    by = "species") %>%
  assert(not_na, join_by, species) %>%
  assert(is_uniq, species) %>%
  tabyl(join_by) %>%
  mutate(join_by = fct_reorder(join_by, n)) %>%
  mutate(
    percent = percent(percent, accuracy = 0.1),
    n = number(n, accuracy = 1)) %>%
  split(.$join_by)

# Plastome ----
# Don't include OG in calculations
n_plast_accs_query <-
  plastome_metadata_raw_all %>%
  left_join(
    select(plastome_metadata_renamed, accession, outgroup),
    by = "accession") %>%
  # excluded accessions lack outgroup info, but are all ferns
  mutate(outgroup = replace_na(FALSE)) %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
n_plast_first_filter <-
  plastome_metadata_renamed %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
n_accs_plast_final <- 
  plastome_seqs_combined_filtered %>%
  left_join(
    select(plastome_metadata_renamed, accession, outgroup),
    by = "accession") %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
```

The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions (note that in some cases a single accession may contain multiple loci).
Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
After excluding these and species names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species (not including species in the plastome dataset).
The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`); `r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).

The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions.
`r n_plast_first_filter` accessions were retained after excluding rogues and species names that could not be resolved.
Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species; `r s_table("accs")`).

@@

Taxon sampling for the current version of FTOL includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`). Taxon sampling for the backbone phylogeny (plastome loci) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.

The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
`r sp_mult_loci$true$n` species (`r sp_mult_loci$true$percent`) were sampled for more than one locus.
Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species).
`r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.

@@

## Topology

```{r topology-stats}
ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))

# Format monophyly status as list for MS
fern_monophy_summ <-
  fern_monophy_summ_tbl %>%
  mutate(
    n = as.integer(n),
    p = percent(percent)
  ) %>%
  split(.$tax_level) %>%
  map(~split(., .$monophyly))

@@

Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (`r figure("backbone")`).
The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
Within leptosporangiate ferns, Osmundales is sister to all other species.
The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales.

The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).

Within Polypodiineae, Hypodematiaceae is sister to all remaining species, which diverge successfully in the order Didymochlaenaceae, Dryopteridaceae, Lomariopsidaceae, Nephrolepidaceae, Tectariaceae, Oleandraceae, then finally Polypodiaceae sister to Davalliaceae.

Within Aspleniineae, Cystopteridaceae is sister to the remaining species, which are split into two clades. 
One clade consists of Rhacidosoraceae, Diplaziopsidaceae, Desmophlebiaceae, Hemidictyaceae and Aspleniaceae.
The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and Athyriaceae.

@@

The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
We recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
This contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have recovered Equisetidae as sister to all other ferns; some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae with low support [@Pryer2001; @Qiu2006; @Qiu2007].
Aside from alternative resolutions of Equisetidae dependent on nuclear envelope or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is lacking.

Another enigmatic relationship in ferns is the placement of Hymenophyllales.
The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates [@Pryer2001; @Schneider2004d; @Schuettpelz2007; @Kuo2011; @Testo2016a].
However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018].
We recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
However, current plastome sampling only includes `r plastome_gleich_genera_n` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.

@@

@@

We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).

@@

Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions (`r figure("genbank")`).
Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
However, the plastid tree does not necessarily agree with trees inferred from other data sources.
Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing, and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
Therefore, we have no expectation that there is a single fern tree of life; rather, multiple trees exist that each capture an element of biological reality, and FTOL is merely one such possible tree.

>>>>

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    sanger_sampling,
    coverage_by_rank,
    total_coverage,
    bb_coverage_by_rank,
    bb_total_coverage,
    gb_species_by_year,
    ref_files,
    other_dates,
    crown_ages,
    fossil_ferns_raw,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    plastome_iqtree_report,
    plastome_iqtree_log
  ),
  store = here::here("_targets")
)
```

@@

We manually inspected any fuzzy matched or non-matching names.
This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
We thus updated and edited the initial World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.

@@

We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
Node support was assessed with up to `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013; @Hoang2018].

@@

We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
We selected `r n_calibrations_ferns` fern fossils to use as constraints from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil). 
We assigned fossils to lineages only after consulting the original publication (as opposed to simply relying on the fossil name) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained for a given fossil could change.
We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
All constraints other than the root are minimum ages.

@@

The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
Code used to generate FTOL and compile this manuscript is available at https://github.com/fernphy/ftol and https://github.com/fernphy/ftol_ms, respectively.
A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.

@@

# Count number of raw accessions in final Sanger selection
# (excluding plastome)
n_accs_raw_final <- count_accs(sanger_accessions_selection_long)

# Count number of accessions in final Sanger selection, split by locus
# (excluding plastome)
n_accs_final <- 
  sanger_accessions_selection_long %>%
  # Count accessions containing different loci as distinct
  mutate(
    name = str_remove_all(name, "accession_"),
    accession = paste(name, accession, sep = "_")) %>%
  count_accs()

# Count number of species in final Sanger selection
# (excluding plastome)
n_species_sanger <- 
  sanger_accessions_selection %>% 
  # Remove outgroup and plastome species
  anti_join(
    unique(select(plastome_seqs_combined_filtered, species)),
    by = "species") %>%
  pull(species) %>%
  n_distinct() %>%
  number()

# Analyze join by method
# (excluding plastome)
join_by_count <-
  sanger_accessions_selection %>% 
  # Remove outgroup and plastome species
  anti_join(
    unique(select(plastome_seqs_combined_filtered, species)),
    by = "species") %>%
  assert(not_na, join_by, species) %>%
  assert(is_uniq, species) %>%
  tabyl(join_by) %>%
  mutate(join_by = fct_reorder(join_by, n)) %>%
  mutate(
    percent = percent(percent, accuracy = 0.1),
    n = number(n, accuracy = 1)) %>%
  split(.$join_by)

# Plastome ----
# Don't include OG in calculations
n_plast_accs_query <-
  # initial search includes some accessions that were later excluded
  plastome_metadata_raw_all %>%
  left_join(
    select(plastome_metadata_renamed, accession, outgroup),
    by = "accession") %>%
  # excluded accessions lack outgroup info, but are all ferns
  mutate(outgroup = replace_na(FALSE)) %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
n_plast_first_filter <-
  plastome_metadata_renamed %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
n_accs_plast_final <- 
  plastome_seqs_combined_filtered %>%
  left_join(
    select(plastome_metadata_renamed, accession, outgroup),
    by = "accession") %>%
  assert(not_na, outgroup) %>%
  filter(outgroup == FALSE) %>%
  count_accs()
```

The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions (note that in some cases a single accession may contain multiple loci).
Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
After excluding these and species names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) after excluding species in the plastome dataset included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species.
The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`); `r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).

The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions, of which `r n_plast_first_filter` accessions were retained after excluding rogues and species with names that could not be resolved.
Selection of the longest set of concatenated sequences per species yielded `r n_accs_plast_final` accessions representing unique species (`r s_table("accs")`).

@@

Taxon sampling for FTOL v1.0.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat v1.0.0).
Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`). Taxon sampling for the backbone phylogeny (plastome loci) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.

The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
A majority of species (`r sp_mult_loci$true$n`; `r sp_mult_loci$true$percent`) were sampled for more than one locus.
Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors are standard deviations unless otherwise mentioned).
The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species); `r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.

@@

## Phylogeny

```{r phylogeny-stats}
# Bootstrap values for selected clades
ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))

# Format monophyly status as list for MS
fern_monophy_summ <-
  fern_monophy_summ_tbl %>%
  mutate(
    n = as.integer(n),
    p = percent(percent, accuracy = 1)
  ) %>%
  split(.$tax_level) %>%
  map(~split(., .$monophyly))

@@

The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_report)`.
The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)`.

Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (`r figure("backbone")`).
The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
Within leptosporangiate ferns, Osmundales is sister to all other species.
The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales. 
The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).

@@

The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
Here, we recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with many other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
However, this contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have all recovered Equisetidae as sister to all other ferns.
Some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae, albeit with generally low support [@Pryer2001; @Qiu2006; @Qiu2007].
Aside from alternative resolutions of Equisetidae dependent on genomic compartment or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is lacking.

Another enigmatic relationship in ferns is the placement of Hymenophyllales.
The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates [@Pryer2001; @Schneider2004d; @Schuettpelz2007; @Kuo2011; @Testo2016a].
However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining (non-Osmundales) leptosporangiates [@Pryer2004a; @Lehtonen2017].
Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018].
Here, we recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
However, current plastome sampling only includes `r plastome_gleich_genera_n` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.

@@

Relationships of families within Aspleniineae (eupolypods II) have been difficult resolve due to the ancient, rapid radiation of this clade [@Rothfels2012].
Our analysis robustly resolves the relationships between all families in Aspleniineae and is in agreement with a recent plastome analysis with similar sampling [@Du2021].
Notably, previous phylogenomic studies that had different or less well-supported topologies did not sample all eupolypod II families [Desmophlebiaceae and Hemidictyaceae absent\; @Wei2017a; @Qi2018; @Shen2018].
The family sister to the remainder of Aspleniineae has been resolved as either Aspleniaceae [`r eg`, @Schneider2004d; @Testo2016a; @Shen2018] or Cystopteridaceae [`r eg`, @Kuo2011; @Wei2017a; @Qi2018].
Here, we recovered Cystopteridaceae as sister to the remainder of eupolypod II families, which are split into two clades.
One clade consists of Rhachidosoraceae, Diplaziopsidaceae, Aspleniaceae, Desmophlebiaceae, and Hemidictyaceae [RHADD clade of @Du2021; Clade E of @Sundue2013].
The other clade includes Thelypteridaceae, Woodsiaceae, Athyriaceae, Onocleaceae, and Blechnaceae [WOBAT clade of @Du2021; Clade B of @Sundue2013].
Each of these two clades is supported by morphological synapomorphies [@Sundue2013; @Du2021].

@@

We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
This may be due to our use of a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).

@@

Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
However, a tree derived from plastid data may not reflect those inferred from other data sources.
Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing [`r eg`, @Sessa2012b; @Zhou2017a; @Wei2021a] and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels. 
Therefore, a major future research goal for fern molecular systematics should be to combine nuclear and plastid datasets to infer species trees with comprehensive sampling.

Recent transcriptomic studies are gradually clarifying the backbone of the fern phylogeny using many (25--`r number(2400)`) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFlag) project, which seeks to generate genomic data (ca. 300 single-to-low copy nuclear gene regions) for all flagellate plants [bryophytes, lycophytes, ferns, and gymnosperms\; @Breinholt2021].
GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.

However, the rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets is still far less than that available for plastid data (`r figure("genbank")`).
Furthermore, many subclades of ferns are under active investigation using both Sanger and next-gen sequencing of plastid markers, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for many years to come.
